---
title: "Human and Economic Costs of Severe Weather in the U.S.A."
author: "martin sneath"
date: "December, 2014"
output: html_document
---
# Synopsis (max 10 sentences)
### Based on U.S.A. event data from 1950 to late 2011 we will analyse fatalities, injury, crop damage and property damage resulting from storm events
### The purpose of this report is to summarize the data in such a way that someone preparing for severe weather events could have some information based on which they could attempt to prioritize their response
# Data Processing
```{r read and preprocess data, cache=TRUE}
# setwd("documents/wp/r/repres/p2")
a=read.csv("repdata-data-StormData.csv.bz2",stringsAsFactors=F)
# There is a cost factor in the data which is a multiplier of the $ field by hundreds, thousands, millions and billions
# there are a minimal number of other factors which are not documented that are assumed to be typos and therefore do not change the dollar value
# multiply by h/k/m/b cost factors
dmgmult=function(arg){
     DMG=as.numeric(arg[1]);mult=arg[2]
     x=1
     if (mult =="h") {x=100}
     else if (mult =="k") {x=1000}
      else if (mult =="K") {x=1000}
      else if (mult =="m") {x=1000000}
      else if (mult =="M") {x=1000000}
else if (mult =="B") {x=1000000000}
     DMG=DMG*x/1000
# multiply dmg by factor, then divide by 1000 to make data presentation easier
     return (DMG)
}
# following applies above function to crop and property damage
a$PDMG=apply(a[c("PROPDMG", "PROPDMGEXP")], 1, FUN=dmgmult)
a$CDMG=apply(a[c("CROPDMG", "CROPDMGEXP")], 1, FUN=dmgmult)
# following makes the date field accessible to date arithmetic
a$YEAR=as.Date(as.character(a$BGN_DATE),"%m/%d/%Y %H:%M:%S")
# there are 985 eventypes in the data set many of which are extremely similar. The following code tries to summarize similar events to make the analysis more comprehensive.
# make eventtype all same case 
a$EVTYPE=tolower(a$EVTYPE)
# remove leading spaces
a$EVTYPE=sub("^\\s+","",a$EVTYPE)
# remove slashes
a$EVTYPE=sub("/"," ",a$EVTYPE)
# reduce variation in event types
a$EVTYPE[grep("freezing rain",a$EVTYPE)]="FREEZING RAIN"
a$EVTYPE[grep("tstm",a$EVTYPE)]="THUNDERSTORM"
a$EVTYPE[grep("thunderstorm",a$EVTYPE)]="THUNDERSTORM"
a$EVTYPE[grep("gustnado",a$EVTYPE)]="THUNDERSTORM"
a$EVTYPE[grep("burst",a$EVTYPE)]="THUNDERSTORM"
a$EVTYPE[grep("rain",a$EVTYPE)]="RAIN"
a$EVTYPE[grep("precip",a$EVTYPE)]="RAIN"
a$EVTYPE[grep("ice",a$EVTYPE)]="ICE"
a$EVTYPE[grep("icy",a$EVTYPE)]="ICE"
a$EVTYPE[grep("glaze",a$EVTYPE)]="ICE"
a$EVTYPE[grep("hail",a$EVTYPE)]="HAIL"
a$EVTYPE[grep("fog",a$EVTYPE)]="FOG"
a$EVTYPE[grep("fire",a$EVTYPE)]="FIRE"
a$EVTYPE[grep("flood",a$EVTYPE)]="FLOOD"
a$EVTYPE[grep("urban sml stream fld",a$EVTYPE)]="FLOOD"
a$EVTYPE[grep("rapidly rising water",a$EVTYPE)]="FLOOD"
a$EVTYPE[grep("wind",a$EVTYPE)]="WIND"
a$EVTYPE[grep("slide",a$EVTYPE)]="SLIDE"
a$EVTYPE[grep("slump",a$EVTYPE)]="SLIDE"
a$EVTYPE[grep("freez",a$EVTYPE)]="FREEZE"
a$EVTYPE[grep("frost",a$EVTYPE)]="FREEZE"
a$EVTYPE[grep("snow",a$EVTYPE)]="SNOW"
a$EVTYPE[grep("blizzard",a$EVTYPE)]="SNOW"
a$EVTYPE[grep("heavy mix",a$EVTYPE)]="SNOW"
a$EVTYPE[grep("sleet",a$EVTYPE)]="SNOW"
a$EVTYPE[grep("dust",a$EVTYPE)]="DUST"
a$EVTYPE[grep("heat",a$EVTYPE)]="HEAT"
a$EVTYPE[grep("warm",a$EVTYPE)]="HEAT"
a$EVTYPE[grep("cold",a$EVTYPE)]="COLD"
a$EVTYPE[grep("hypothermia",a$EVTYPE)]="COLD"
a$EVTYPE[grep("low temperature",a$EVTYPE)]="COLD"
a$EVTYPE[grep("tide",a$EVTYPE)]="TIDE"
a$EVTYPE[grep("tornado",a$EVTYPE)]="TORNADO"
a$EVTYPE[grep("landspout",a$EVTYPE)]="TORNADO"
a$EVTYPE[grep("funnel",a$EVTYPE)]="TORNADO"
a$EVTYPE[grep("lightning",a$EVTYPE)]="LIGHTNING"
a$EVTYPE[grep("hurricane",a$EVTYPE)]="HURRICANE"
a$EVTYPE[grep("tropical",a$EVTYPE)]="TROPICAL STORM"
a$EVTYPE[grep("sea",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("swell",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("surf",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("waves",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("rip",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("erosion",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("surge",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("high",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("marine",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("rogue wave",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("coastal",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("drowning",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("seiche",a$EVTYPE)]="SEAS&SURF"
a$EVTYPE[grep("wint",a$EVTYPE)]="WINTER"
# fix known typos in data
a$EVTYPE[grep("avalance",a$EVTYPE)]="avalanche"
a$EVTYPE[grep("water spout-",a$EVTYPE)]="water spout"
a$EVTYPE[grep("hyperthermia",a$EVTYPE)]="COLD"
# Aggregate processed data by summarized type
costs=aggregate(cbind(a$FATALITIES,a$INJURIES,a$PDMG,a$CDMG),by=(list(a$EVTYPE)),FUN=sum)
colnames(costs)=c("EVTYPE","FATALITIES","INJURIES","PDMG","CDMG")
```
# Results - Tornadoes are number 1 source of fatalities and injuries
## The data seems to be overreported for damage and events in the late naughties while fatality and Injury figures seem to be reported more   consistently.
### Flood is the major cause of property damage, while drought is the number 1 cause of crop damage, but followed very closely by the opposite condition of drought

```{r results}
# analyze data by year
c=substr(a$YEAR,1,4)
par(mfrow=c(2,2))
hist(as.numeric(c),main="Analyis by Year",xlab="",ylab="Number of Events")
plot(c,(a$PDMG),type="h",ylab="$ Damage")
plot(c,a$FATALITIES,type="h",ylab="Fatalities")
plot(c,a$INJURIES,type="h",ylab="Injuries")
# analyze top 10 events in each category of loss/damage - uses dplyr arrange
library(dplyr,quietly)
par (las=2,mfrow=c(1,2))
f=arrange(costs,desc(FATALITIES))
i=arrange(costs,desc(INJURIES))
p=arrange(costs,desc(PDMG))
c=arrange(costs,desc(CDMG))
f=f[1:10,];i=i[1:10,];p=p[1:10,];c=c[1:10,]   
barplot(f$FATALITIES,horiz=T,names.arg=f$EVTYPE,cex.names=.5,main="Fatalities")
barplot(i$INJURIES,horiz=T,names.arg=i$EVTYPE,cex.names=.5,main="Injuries")
# Total Fatalities Reported 1950 - 2011
sum(costs$FATALITIES)
# Total Injuries Reported 1950 - 2011
sum(costs$INJURIES)
par (las=2,mfrow=c(1,2))
# following stops $ being reported in scientific notation
options(scipen=999)
barplot(p$PDMG,horiz=T,names.arg=p$EVTYPE,cex.names=.5,main="Prop Damage K $")
barplot(c$CDMG,horiz=T,names.arg=c$EVTYPE,cex.names=.5,main="Crop Damage K $")
# total costs property & crop damage 1,000s of $
sum(a$PDMG,a$CDMG)
# crop damage as a % of property damage
sum(a$CDMG)/sum(a$PDMG)*100
```
